---
title: "Using tinyForestData"
format: html
editor: visual
execute: 
  message: false
  warning: false
  cache: false
---

## Introduction

`tinyForestData` is an R package which accompanies the `tinyForestR` package. It contains datasets which allow exploration of the development and biodiversity associated with Tiny Forests in the UK. It currently contains x datasets.

### `tf_latest_sf_uk`

This a simple features dataset using EPSG 4326 as coordinate reference system which contains data on 196 UK Tiny Forests including:

-   Name and ID

-   Website

-   Planting date

-   Planted area

-   Mix of trees

-   Age in days (from planted date to July 1st 2023)

-   xy coordinates as a `geometry` field

-   Using the `sf` format enables easy mapping as well as data manipulation

```{r, echo=FALSE}
library(sf); library(mapview); library(tinyForestData); library(tidyverse)



tf_latest_sf_uk |>
  mapview()




```

### 

### `tf_urban_classification`

This includes the ONS (and Scottish equivalent) urban rural classification of the location of each TF.

```{r}

tf_urban_classification |> 
  count(UR_class) |>
  ggplot(aes(reorder(UR_class, n), n)) +
  geom_col() +
  coord_flip()
  
```

### `tf_nbn_df_2010`

This is a large dataset (\> 500000 rows of data) with observations of species within a 1km buffer of each TF since 2010 derived from the National Biodiversity Network API.

```{r}

tf_nbn_df_2010 |>
  group_by(tfId, year) |>
  count() |>
  ggplot(aes(year, fct_rev(factor(tfId)), fill = n)) +
  geom_tile() +
  scale_fill_viridis_c(option = "turbo")


```

### landcover and greenspace

```{r}

library(sf)

geo_tf <- geo_tf |> st_as_sf()

geo_tf <- geo_tf |>
  mutate(area = st_area(geo_tf))
geo_tf_dt <- data.table::data.table(geo_tf)

geo_tf_dt[, .(n = .N, s_area = sum(area)), by = .(tf_id, dscrptn)][] |>
  data.table::fwrite("data/tf_lc.csv")
  
```
